---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# needed libraries
library(tidyr)
library(dplyr)
library(ggplot2)
library(gtools)
library(pROC)
library(stringr)
# source from file the combiroc custom functions
source("C:/Users/rossiriccardo/R/combiroc/combiroc_functions.R")
```

## Data loading

Load the proteomic dataset from Zingaretti et al. 2012. Dataset should be formatted in a specific way. The load() function uses a customized read.table() function and performs a number of checks on the content of the loaded datasets.
The loaded dataset is assigned to the "data" object.

```{r}
data <- load("data/demo_5Ags.csv", sep=';')
head(data)
data_long <- CombiROC_long(data) # to make data in long format
```

## Markers distribution overview

To plot boxplot with selecting the y value lim (zoom only, no data loss) and prints summary for each class

```{r}
markers_overview(data_long, ylim =2000) 
```

